---
title: "Using rapid temperature falls to estimate future strong cold front frequency in CMIP6 climate projections"
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
    docx: default
author:
  - name: Paola B Corrales
    affiliations:
      - name: Monash University
        department: School of Earth, Atmosphere and Environment
        region: Victoria
        country: Australia
    corresponding: true
    orcid: 0000-0003-1923-9129
    email: fake_email@fakeyfake.com
  - name: Martin Singh
    affiliations:
      - name: Monash University
        department: School of Earth, Atmosphere and Environment
        region: Victoria
        country: Australia
    orcid: 0000-0001-8715-9476
    email: martin.singh@monash.edu
  - name: Michael J. Reeder
    affiliations:
      - name: Monash University
        department: School of Earth, Atmosphere and Environment
        region: Victoria
        country: Australia
    orcid: 0000-0001-8715-check
    email: fake_email@fakeyfake.com
abstract: |
  We use day-to-day falls in surface maximum temperature to assess future changes in cold front frequency in a set of CMIP6 climate models under the SSP5-8.5 scenario. We find a general decrease in cold front frequency across the Northern Hemisphere, particularly over the oceans, and a poleward shift in the Southern Hemisphere, with increased frequency over midlatitude land regions. The frequency of strong temperature falls correlates positively with the meridional temperature gradient and transient eddy kinetic energy, highlighting the role of large-scale baroclinicity and storm track changes in future frontal behavior. Over Southern Hemisphere land areas, this relationship weakens, suggesting the influence of regional thermodynamic processes.
plain-language-summary: |
  Cold fronts, weather systems associated with large temperature falls and sometimes rain and strong wind, play a key role in the day-to-day weather in midlatitudes. In this study, we use a simple method based on changes in maximun temperature from one day to the next to identify cold fronts in climate simulations. Looking at a future climate projection of ten global models, we find that cold fronts are expected to become less frequent in much of the Northern Hemisphere, especially over the oceans. In the Southern Hemisphere, however, cold fronts are projected to shift closer to the South Pole and may become more frequent over land areas like southern Australia and southern Africa. These changes are linked to shifts in the temperature contrast between the equator and the poles and the paths taken by storm systems. 
keywords: []
keypoints: 
  - Day-to-day falls in maximun temperature are used to identify for cold fronts.  
  - CMIP6 SSP5-8.5 scenario show a decrease in frequency of falls in temperature in the Northern Hemisphere, and a poleward shift over the oceans and an increase over land in the Southern Hemisphere.
  - Changes in cold front frequency are multicausal and could be related to changes in the meridional gradient of surface temperature and storm tracks among other processes.
bibliography: [cold-fronts.bib, packages.bib]  
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
execute: 
  cache: true
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup

library(metR)
library(ggplot2)
library(data.table)
# library(lubridate)
library(here)
library(cowplot)
library(ggside)
library(patchwork)

global_map <- rnaturalearth::ne_coastline()
models_in_gadi <- c("CMCC-ESM2", "EC-Earth3", "INM-CM4-8", "INM-CM5-0", 
                    "MPI-ESM1-2-HR", "NorESM2-MM", "EC-Earth3-CC", "EC-Earth3-Veg",
                    "EC-Earth3-Veg-LR", "GFDL-CM4")

file_list <- c(Sys.glob(here("data/eke/*hist*mean.nc")),
               Sys.glob(here("data/deltat/*hist*mean.nc")),
               Sys.glob(here("data/dQ-dy/*hist*mean.nc")))

delta <- purrr::map(file_list, function(f) {
  
  message(f)
  meta <- unglue::unglue(basename(f), "{var}_{model}_{scenario}_mean.nc")
  
  
  if (meta[[1]][["var"]] == "eke") {
    var <- "ua"
  } else if (meta[[1]][["var"]] == "deltap97.5") {
    var <- "tasmax"
    meta[[1]][["var"]] <- "deltat"
  } else {
    var <- "hfss"
  }
  
  file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/*_", meta[[1]][["model"]], "_ssp585_mean.nc")))
  
  ReadNetCDF(f, vars = c(hist = var)) |> 
    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
    _[, let(model = meta[[1]][["model"]],
            var = meta[[1]][["var"]],
            change = ssp585 / hist,
            diff = ssp585 - hist,
            lat = round(lat, digits = 8),
            plev = NULL)] |> 
    _[]
  
}) |> rbindlist() 


ta <- purrr::map(Sys.glob(here("data/ta/*hist*mean.nc")), function(f) {
  
  message(f)
  meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_historical_mean.nc"))
  
  file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/", 
                                        meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585*")))
  
  var <- meta[[1]][["var"]]
  
  ReadNetCDF(f, vars = c(hist = var), subset = list(plev = 85000)) |> 
    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), 
                                subset = list(plev = 85000),
                                out = "vector")[[1]])] |> 
    _[, let(model = meta[[1]][["model"]],
            var = var,
            lat = round(lat, digits = 8),
            hist = metR:::.derv(hist, lat)*(180/pi/6371000),
            ssp585 = metR:::.derv(ssp585, lat)*(180/pi/6371000)), by = .(lon)] |> 
    _[, let(change = ssp585 / hist,
            diff = ssp585 - hist,
            plev = NULL)]
  
}) |> rbindlist() 

data <- rbindlist(list(delta, ta)) |> 
  _[, let(land = MaskLand(lon, lat))] |>
  _[, let(land = factor(land, labels = c("Sea", "Land")))]

```



## Introduction

Cold fronts are an important feature of weather in extratropical regions and their associated circulation contribute a substantial fraction of the preciptiation in these regions [@Konstali2024]. Cold fronts are also linked to extreme events, such as heavy precipitation [@Catto2013] and severe wildfires in southern Australia [@Reeder2015] and other regions. Understanding how cold fronts change with warming is therefore of importance for understanding the changing likelihood of extremes and fire weather.

Several computational methods have been developed to identify cold fronts using different characteristics. Some of these methods look for thermodynamic changes associated with temperature and humidity at different atmospheric levels [@Berry2011], others identify changes in the wind field alone [@Simmonds2012]. These methods rely on the data being in a regular grid (e.g. outputs from numerical models or reanalysis) and their results depend on setting different thresholds, which need to be calibrated for different grid resolutions. 


Here, we take a simpler approach, first introduced by @Reeder2015. We measure the occurrence of cold fronts using temporal changes in temperature at a fixed location. Specifically, we identify the passage of cold fronts with large day-to-day falls in near-surface daily-maximum temperature. This method has been shown to have success across different resolutions, including both in station observations and reanalyses [@King2024]. Previous studies have shown that this method is suitable for the detection of strong cold fronts in southeastern Australia [@Cai2022; @Reeder2015], and recently, @King2024 demonstrated that this method can successfully identify cold fronts with results comparable to those obtained with other methodologies. Furthermore, @King2024 calculated a climatology of front frequency using global climate models from phase 6 of the Coupled Model Intercomparison Project (CMIP6). While the models tend to overestimate front frequency in polar regions and underestimate the occurrence of large temperature falls over midlatitude land regions compared with ERA5, they accurately represent the patterns of front frequency equatorward of about 60°.

As global temperatures continue to rise, understanding how this warming will impact the occurrence of cold fronts is crucial for projecting changes in extreme weather and fire risk. One of the few studies to have addressed this question is @Catto2014 who used objective front identification using temperature and humidity fields [@Berry2011] in 18 CMIP5 models and compared the changes between historical simulations between 1980-2005 and the Representative Concentration Pathway 8.5 between 2080-2100. They found an overall decrease in front frequency in the Northern Hemisphere, with a poleward shift in the maximum frequency and a significant decrease at high latitudes. The Southern Hemisphere showed a poleward shift in the frequency maximum. On the other hand, @Konstali2024, identified fronts in the CESM2-LE model simulations using the gradient of equivalent potential temperature in 850 hPa and found an increase in the frequency of fronts that could be related to fronts becoming larger or more elongated in both hemispheres. 

Motivated by its simplicity and robustness to resolution, here we evaluate future changes in cold front frequency based on day-to-day falls in temperature across the CMIP6 ensemble. This approach offers several advantages. First, the method's simplicity and resolution-independence allows us to compare it with earlier studies using other methodologies. Second, focusing on local temperature falls provides a complementary perspective to Lagrangian approaches, as day-to-day temperature falls are the most direct impact of cold fronts on local weather.
Consistent with previous work, we find an increase in cold front frequency in the northern hemisphere and a poleward shift in the Southern Hemisphere. Seeking to develop a physical explanation for these changes, we first studied the response of the meridional temperature gradient and it relationship with the change in front frequency. We also explored the relationship between changes in front frequency and transient eddy kinetic energy, as it is directly related to the storm tracks. Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@OGorman2010; @Tamarin2017; @Chemke2022], and these changes could also influence the location and intensity of future cold fronts.


## Data and methods

We analysis 10 models from phase 6 of the Coupled Model Intercomparison Project (CMIP6, @Eyring2016), including the Historical experiment up to 2014 and the Shared Socioeconomic Pathways 5 (SSP5-8.5) from the Scenario Model Intercomparison Project (ScenarioMIP, @ONeill2016) as described in @tbl-models. We selected these models based on the data available for the  historical and scenario experiments stored on the Australian National Computational Infrastructure (NCI) replica data store.


```{r}
#| label: tbl-models
#| tbl-cap: "List of the coupled model intercomparison project phase 6 (CMIP6) climate model used in this analysis."
#| results: asis


readr::read_rds(here("data/cmip_file_list_scenario.rds")) |> 
  _[nominal_resolution == "100 km" & source_id %in% models_in_gadi, 
    .(source_id, institution_id, nominal_resolution)] |> 
  unique(by = "source_id") |> 
  _[order(source_id)] |> 
  _[, let(references = c("@Lovato2021; -@Lovato2021a", "@ConsortiumEC-Earth2019; -@ConsortiumEC-Earth2019a",
                         "@ConsortiumEC-Earth2020; -@ConsortiumEC-Earth2021", 
                         "@ConsortiumEC-Earth2019b; -@ConsortiumEC-Earth2019c",
                         "@ConsortiumEC-Earth2020a; -@ConsortiumEC-Earth2023",
                         "@Guo2018; -@Guo2018a", "@Volodin2019, @Volodin2019a",
                         "@Volodin2019b; @Volodin2019c", "@Jungclaus2019; @Schupfner2019",
                         "@Bentsen2019; @Bentsen2019a"))] |> 
  _[, let(source_id = paste0(source_id, "\n[", references, "]"))] |> 
  _[, .(source_id, institution_id, nominal_resolution)] |> 
  knitr::kable(col.names = c("Model", "Institution", "Original resolution"),
               caption = "List of models used for the analysis.",
               align = "llc",
               format = "markdown", format.args = list(zero.print = "-"))

```


### Front Identification

We identify falls in surface air temperature as a proxy for frontal passages by calculating the change in daily maximum temperature from one day to the next ($\Delta T$). $\Delta T$ is calculated for all grid points and days in the study period. @King2024 defined a passage of a strong cold front as a $\Delta T \le -10^{\circ}C$. That is, a cold front passage is identified on day $i$ if the maximum temperature on day $i+1$ is at least 10 deg C lower than that of day $i$. While this value is reasonable for detecting strong falls in temperature over land, it may underestimate the frequency of fronts over the ocean due to the smaller temperature variability. For this reason, in this study, we use the value of the 2.5th percentile of $\Delta T$ (as we want to capture the most negative temperature falls) as a threshold for each grid point to identify falls in temperature, i.e. $\Delta T \le 2.5^{th}$ percentile. This percentile is chosen because the mean of the 2.5th percentile over all midlatitude land areas (between 25$^{\circ}$ to 55$^{\circ}$) is close to $-10^{\circ}C$, the threshold used by @King2024. The percentiles are calculated over the historical period between 1979 and 2014 for each model at each grid point. It is important to note that by using this threshold we detect frontal passages in exactly 2.5% of days. In regions of regular frontal passage the 2.5th percentile of $\Delta T$ generally corresponds to strong fronts, and our identification method produces reasonable results. However, outside these regions (e.g., in the tropics) we identify temperature drops that can be associated to midlatitude frontlike features but also other processes. 



we detect falls in temperature in the tropics that are unrelated to frontal systems. For this reason, the analysis will focus only on mid and high latitudes.


### Time-mean meridional temperature gradient

We calculated the time-mean meridional temperature gradient ($\partial \bar{T}/ \partial y$) at each grid point using central finite differences for the temperature at 850 hPa. We then area-averaged $\partial \bar{T}/ \partial y$ over different regions, i.e.  between 25$^{\circ}$ and 65$^{\circ}$ latitude for each hemisphere. This process produces the same results as if calculating the temperature difference between two latitudes but give us the advantage of calculating the mean $\partial \bar{T}/ \partial y$ for regions over land and sea independently.

### Storm Tracks estimation

To estimate the storm track location we calculated the transient eddy kinetic energy (EKE), used by many others to estimate storm tracks [@OGorman2010; @Chemke2022]. For each model, we calculated the EKE using a Fast Fourier Transform high-pass filter applied to daily horizontal wind at 300 hPa with a 6-day cutoff.  



## Results

@fig-trend a) shows the change in the frequency of $\Delta T$ between the yearly averaged $\Delta T$ frequency for last 20 years of the 21$^{st}$ century (SSP5-8.5 experiment) and
the last 20 years of the 20$^{th}$ century (historical experiment). @fig-trend b and c) show the zonal mean of $\Delta T$ frequency for each latitude over sea and over land respectively. In the Northern Hemisphere over land, there is a general decrease in strong temperature falls, particularly north of 40$^{\circ}$ latitude, except in western and central Europe where there is an increase. A similar pattern of changes in front frequency was also found by @Catto2014 using CMIP5 and a different methodology to identify fronts, reinforcing the reliability of the methodology used in this analysis. Over the oceans, there is also a decrease in the frequency of fronts except over the North Atlantic, which shown an increase in front frequency west of continental Europe. This local increase may be interpreted as a shift of front frequency toward the pole in this sector.

In the Southern Hemisphere over land, the change is generally positive with an increase of 5 strong fronts per year in some areas. Other authors have also found an increase in the frequency of fronts in the south and east regions of Africa and Australia [@Catto2014; @Konstali2024]. 

Over oceanic regions, the change in frequency of frontal passages shows a tripolar pattern, with increases in a band centered around 40$^{\circ}$-50$^{\circ}$ S (depending on the longitude) and decreases poleward and equatorward. 
This pattern indicates that the areas of higher front frequency shift poleward in the future scenario. In the Indian ocean, the pattern in more compicated, with a second band of increased frequency in the subtropics most pronounced in the Agulas region. The oceanic regions in the Southern Hemisphere is also associated with an increase in $\partial \bar{T}/ \partial y$ and in storm track activity (not shown).


```{r}
#| label: fig-trend
#| fig-width: 8
#| fig-height: 4
#| fig.cap: "a) Multimodel average change in the frequency of $\\Delta T$ (days/year, SSP5-8.5 2018-2100 minus Historical 1980-2000) and zonal mean change over sea b) and land c), each grey line represents the models listed in Table 1 and in black, the multimodel mean. In a) high topography areas are gray-shaded and dotted areas show where at least 90% of the models agree in the sign of change."

topo <- rcdo::cdo_topo(grid = "r360x180") |> 
  rcdo::cdo_execute(options = "-f nc") |> 
  ReadNetCDF()

sig <- data |> 
  _[(model %in% models_in_gadi) & var == "deltat"] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, .(sig = sum(sign(deltat), na.rm = TRUE)), by = .(lon, lat)] 

map_trend <- data |> 
  _[(model %in% models_in_gadi) & is.finite(change)] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, .(mean_dt = mean(deltat, na.rm = TRUE)), by = .(lon, lat)] |> 
  ggplot(aes(ConvertLongitude(lon), lat)) +
  geom_contour_fill(aes(z = mean_dt, fill = after_stat(level)),
                    breaks = c(-Inf, seq(-8, 8, 0.5), Inf)) +
  scale_fill_divergent_discretised(name = "days/year", mid = "#f7f7f7",
                                   guide = guide_colorbar(barheight = 0.5),
                                   labels = function(x) JumpBy(x, 2, fill = "")) +
  geom_sf(data = global_map, inherit.aes = FALSE, fill = NA, linewidth = 0.2) +
  geom_point(data = sig[is.cross(lon, lat, skip = 1) & abs(sig) >= 9], alpha = 0.3, shape = 4, size = 0.1) +
  ggnewscale::new_scale_fill() +
  geom_contour_fill(data = topo, aes(ConvertLongitude(lon), lat, z = topo), breaks = c(1300, Inf), fill = "grey90") +
  coord_sf(expand = FALSE, xlim = c(-179, 180)) +
  labs(x = NULL, y = NULL, fill = "days/year") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.frame = ggplot2::element_rect(color = "black", linewidth = 0.4),
        legend.key.width = grid::unit(1, 'null'),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), 
        plot.margin = unit(c(0, -0.4 , 0 , 0), units = "cm")) 

annotation <- data.frame(ymin = -66, ymax = -56, xmin = -Inf, xmax = Inf, land = "Land")

mean_trend <- data |> 
  _[(model %in% models_in_gadi) & var == "deltat"] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, .(mean_dt = mean(deltat, na.rm = TRUE)), by = .(lat, land, model)] |> 
  ggplot(aes(mean_dt, lat)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
  stat_summary(orientation = "y", geom = "line") +
  scale_y_latitude(breaks = seq(-80, 80 , 20)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#fbfbfb",
            data = annotation,
            inherit.aes = FALSE) +
  facet_wrap(~land) +
  tagger::tag_facets(tag_pool = c("b", "c")) +
  labs(x = "days/year") +
  coord_fixed(ratio = 1/2.8) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        panel.grid.minor = element_line(linewidth = 0.1),
        strip.text.x = element_text(size = 12),
        plot.margin = unit(c(0, 0 , 1.9, -0.40), units = "cm"),
        tagger.panel.tag.text = element_text(size = 12),
        tagger.panel.tag.background = element_rect(fill = "white", color = "white")) 

plot_grid(map_trend, mean_trend, labels = c('a)', ''), 
          label_x = 0, label_y = 0.94,
          label_fontface = "plain", 
          label_colour = "white",
          rel_widths = c(1.75, 1))

```

@fig-serie shows ensemble-mean time series of the zonal-mean frequency of strong temperature drops for different latitude bands in each hemisphere. Consistent with the map in @fig-trend, the trends in frequency are negative over ocean in the Northern hemisphere, but positive over land equatorward of 50$^{\circ}$ N. Over the Southern hemisphere, trends are generally positive over land, with changes over ocean generally consistent with a poleward shift of fronts.

While there is considerable variation between models in these trends (see boxplots on right of @fig-serie), particularly at more poleward locations, for most of the zonal mean trends, there is consistency in the sign of the trend across models.


```{r}
#| label: fig-serie
#| fig-width: 8
#| fig-height: 5
#| fig.cap: "Multimodel mean frequency of $\\Delta T$ for the historical period (1850-2014) and SSP5-8.5 (2015-2100) over land and sea for different 2° wide latitude bands centered at 30, 40 and 50°. Right to each main subfigure the intermodel distribution is shown for the period 2080-2100."

serie <- purrr::map(Sys.glob(here("data/series/serie_freq_deltap97.5_*_100km.nc")), function(f) {
  
  meta <- unglue::unglue(basename(f), "serie_freq_{threshold}_{scenario}_100km.nc")
  ReadNetCDF(f, vars = c(mean_freq = "tasmax")) |> 
    _[, let(year = year(time),
            scenario = meta[[1]][["scenario"]])] |> 
    _[, let(time = NULL)]
  
}) |> rbindlist() |> 
  _[, let(scenario = factor(scenario, labels = c("Historical", "SSP5-8.5")))]

model_spread <- data |> 
  _[var == "deltat" & model %in% models_in_gadi] |> 
  _[, let(serie = fcase(lat %between% c(48, 50), "50° - NH",
                        lat %between% c(38, 40), "40° - NH",
                        lat %between% c(28, 30), "30° - NH", 
                        lat %between% c(-50, -48), "50° - SH",
                        lat %between% c(-40, -38), "40° - SH",
                        lat %between% c(-30, -28), "30° - SH"))] |>
  _[!is.na(serie)] |>
  _[, .(mean_freq = weighted.mean(ssp585, cos(lat*pi/180))), by = .(serie, land, model)] |> 
  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
  setDT() |> 
  _[, let(scenario = "SSP5-8.5")] |> 
  _[!(land == "Land" & hemisphere == "SH" & lat == "50°")]

serie_regions <- serie |>
  _[!(year == 2014 & scenario != "historical")] |>
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, let(serie = fcase(lat %between% c(48, 50), "50° - NH",
                        lat %between% c(38, 40), "40° - NH",
                        lat %between% c(28, 30), "30° - NH", 
                        lat %between% c(-50, -48), "50° - SH",
                        lat %between% c(-40, -38), "40° - SH",
                        lat %between% c(-30, -28), "30° - SH"))] |>
  _[!is.na(serie)] |> 
  _[, .(mean_freq = weighted.mean(mean_freq, cos(lat*pi/180))), by = .(land, serie, year, scenario)] |> 
  _[!(land == "Land" & serie == "50° - SH")]

serie_regions |> 
  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
  setDT() |> 
  _[year >= 1900] |> 
  ggplot(aes(year, mean_freq, color = lat, group = interaction(scenario, lat))) +
  geom_line(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_ysideboxplot(data = model_spread, aes(x = lat), orientation = "x") +
  scale_color_manual(values = c("#004C73", "#CC3311", "#ee8a33")) + 
  facet_grid(land~hemisphere, scale = "free") +
  labs(x = "Year", y = latex2exp::TeX("$\\Delta T$ frequency (days/year)"), color = "Latitude") +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.78, 0.07),
        legend.direction = "horizontal",
        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        strip.text = element_text(size = 12),
        ggside.axis.text.x = element_blank(),
        ggside.panel.scale = 0.2)

```
To explore the distribution of the change in temperature falls of different intensity, @fig-distribution shows the change in frequency of $\Delta T$ using different percentiles, from 10$^{th}$ to 0.1$^{th}$, as thresholds to identify the falls in temperature. The percentile values are again calculated from the historical period between 1979 and 2014 for each grid point and model as in @fig-trend. 
This figure shows that the change in the frequency of $\Delta T$ is greater at lower percentile thresholds, which identify more extreme temperature falls. For example in the Northern Hemisphere over the sea, the 0.1 percentile threshold shows that there will be less than half as many falls in temperature in the future over 50$^{\circ}$ N, with the opposite expected over land in the Southern Hemisphere. 
However, it is important to note that lower percentiles thresholds are associated with rarer events, meaning fewer occurrences per year. These results suggest that regions projected to experience more temperature falls in the future, such as land areas in midlatitudes in the Southern Hemisphere, will also experience more intense $\Delta T$, whereas regions with a decreasing frequency of events will see less intense falls in temperature.


```{r}
#| label: fig-distribution
#| fig-width: 8
#| fig-height: 5
#| fig.cap: "Fractional change in the frequency of $\\Delta T$ (SSP5-8.5 2018-2100 divided Historical 1980-2000) for different thresholds defined as the percentile value for each grid point and model calculated over the historical period (1979-2014)."

file_list <- Sys.glob(here(paste0("data/series/serie_freq_deltap9*", c("historical", "ssp585"), "*")))

dist <- purrr::map(file_list, function(f) {
  message(f)
  
  meta <- unglue::unglue(basename(f), "serie_freq_deltap{percentile}_{scenario}_100km.nc")
  
  if (meta[[1]][["scenario"]] == "historical") {
    
    ReadNetCDF(f, vars = "tasmax", subset = list(time = c("1980-01-01", "2000-12-31"))) |> 
      _[, .(mean_freq = mean(tasmax)), by = .(lon, lat)] |> 
      _[, let(percentile = meta[[1]][["percentile"]],
              scenario = meta[[1]][["scenario"]])]
  } else {
    
    ReadNetCDF(f, vars = "tasmax", subset = list(time = c("2080-01-01", "2100-12-31"))) |> 
      _[, .(mean_freq = mean(tasmax)), by = .(lon, lat)] |> 
      _[, let(percentile = meta[[1]][["percentile"]],
              scenario = meta[[1]][["scenario"]])]
  }
  
}) |> rbindlist() |> 
  _[, let(land = MaskLand(lon, lat),
          region = fcase(lat %between% c(25, 55), "NH",
                         lat %between% c(-55, -25), "SH",
                         default = NA))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] 

dist |> 
  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50° - NH",
                        lat %between% c(38, 40) & land == "Sea", "40° - NH",
                        lat %between% c(28, 30) & land == "Sea", "30° - NH", 
                        lat %between% c(48, 50) & land == "Land", "50° - NH",
                        lat %between% c(38, 40) & land == "Land", "40° - NH",
                        lat %between% c(28, 30) & land == "Land", "30° - NH",
                        lat %between% c(-50, -48) & land == "Land", "50° - SH",
                        lat %between% c(-40, -38) & land == "Land", "40° - SH",
                        lat %between% c(-30, -28) & land == "Land", "30° - SH",
                        lat %between% c(-50, -48) & land == "Sea", "50° - SH",
                        lat %between% c(-40, -38) & land == "Sea", "40° - SH",
                        lat %between% c(-30, -28) & land == "Sea", "30° - SH"))] |>
  _[!is.na(serie)] |> 
  _[, .(mean_freq = weighted.mean(mean_freq, cos(lat*pi/180))), by = .(land, serie, percentile, scenario)] |> 
  dcast(land + serie + percentile ~ scenario, value.var = "mean_freq") |> 
  _[, let(historical = fifelse(serie  == "50° - SH" & land == "Land", 0, historical),
          ssp585 = fifelse(serie  == "50° - SH" & land == "Land", 0, ssp585))] |> 
  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
  setDT() |> 
  ggplot(aes(factor(round(100-as.numeric(percentile), digits = 1)), ssp585 / historical)) +
  geom_col(aes(fill = lat), position = "dodge") +
  scale_fill_manual(values = c("#004C73", "#CC3311", "#ee8a33")) +
  scale_y_log10() +
  facet_grid(land~hemisphere) +
  labs(x = "Percentile", y = latex2exp::TeX("Fractional change of $\\Delta T$"),
       fill = "Latitude") +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.8, 0.07),
        legend.direction = "horizontal",
        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        strip.text = element_text(size = 12))


```

### Mecanisms

We now investigate the mechanisms driving the changes in strong $\Delta T$ frequency described above. We consider a simple perspective on midlatitude temperature variability based on advection as a key driver [@schneider2015]. According to this view, changes in midlatitude temperature variability are closely linked to changes in large-scale temperature gradients and to the strength of mixing by geostrophic turbulence. Motivated by this perspective, we explore relationships between changes in front frequency and changes in the large-scale meridional temperature gradient ($\partial \bar{T} / \partial y$) and the eddy kinetic energy (EKE) within the midlatitude storm tracks.

@fig-eke a) shows that, in both hemispheres, there is a strong correlation between increases in front frequency and changes in the large-scale meridional surface temperature gradient $\partial \bar{T} / \partial y$. Here, $\partial \bar{T} / \partial y$ is taken as the time and zonal-mean meridional temperature gradient averaged between 25$^{\circ}$ and 65$^{\circ}$ in each hemisphere over either land or ocean.
As previously noted by @Catto2014, reductions in front frequency in the Northern Hemisphere are associated with a decrease in the the temperature gradient near the surface, while opposite changes in surface temperature gradients in the Southern Hemisphere drive increases is front frequency there. Here, we find that this relationship holds within the model ensemble; larger increases or decreases in  $\partial \bar{T} / \partial y$ in a given model are associated with larger projected increases or decreases in $\Delta T$. 
A similar relationship across the ensemble is present when comparing changes in the frequency $\Delta T$ and changes in the energy of the storm track. Here we estimate the storm track energy by taking the eddy kinetic energy at 300 hPa, applying a high band-pass filter with a 6-day cutoff, and integrating over the 25$^{\circ}$ and 65$^{\circ}$ latitudes in each hemisphere. Models with projected decreases in EKE tend to project larger decreases in the frequency of strong $\Delta T$. 
The exception is the Southern Hemisphere over land where all models project weak changes in EKE and large increases in the frequency of strong $\Delta T$. The reason for this difference is unclear, but could suggest the importance of local land processes in amplifying fronts over land in some regions of the Southern Hemisphere (cite papers by Michael). Another hypothesis is that the most of the land affected by fronts in the Southern hemisphere is north of the mean storm tracks, and thus the storm tracks has only a weak effect on temperature variability there.
Supporting this, @Harvey2014 found that the storm track response in the Southern Hemisphere during summer is associated with the change in the low level $\partial \bar{T} / \partial y$ mainly poleward of 65$^{\circ}$ S, i.e. over the southern oceans. 


Notwithstanding the above caveat, our results suggest that changes in the mean meridional temperature gradient and storm track intensity play an important role in modulating changes in front frequency in a warming climate. Indeed, differences in how these quantities change in projections appear to account for the differences in front frequency changes between hemispheres. However, our results do not provide strong evidence for the relative importance of the two factors: aside from over Southern Hemisphere land, both factors correlate with the $\Delta T$ frequency across the model ensemble and both are theoretically associated with changes in temperature variability.


```{r}
#| label: fig-eke
#| fig-width: 8
#| fig-height: 4
#| fig.cap: "Mean change in $\\Delta T$ frequency and a) $\\partial \\bar{T} / \\partial y$ and b) EKE for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model. Values are calculated for the northern (25$^{\\circ}$ to 65$^{\\circ}$) and southern (-25$^{\\circ}$ to -65$^{\\circ}$) hemispheres and over sea (circles) and land (triangles). Each solid line corresponds to a linear regression for each hemisphere."

change <- data |> 
  _[(model %in% models_in_gadi)] |>
  dcast(lat + lon + model + land ~ var, value.var = "diff") |> 
  
  _[, let(region = fcase(lat %between% c(25, 65), "NH",
                         lat %between% c(-65, -25), "SH",
                         default = NA))] |> 
  _[, .(mean_eke = weighted.mean(eke, w = cos(lat*pi/180), na.rm = TRUE),
        mean_dt_dy = weighted.mean(ta, w = cos(lat*pi/180), na.rm = TRUE),
        mean_deltat = weighted.mean(deltat, w = cos(lat*pi/180), na.rm = TRUE)), by = .(region, land, model)] |> 
  _[, let(mean_dt_dy = fifelse(region == "NH", -mean_dt_dy, mean_dt_dy))] |> 
  na.omit()

gradt <- change |> 
  ggplot(aes(mean_dt_dy*1000, mean_deltat)) +
  geom_point(aes(color = region, shape = land), size = 2) +
  geom_smooth(aes(colour = region), method = "lm", se = FALSE) +
  scale_color_manual(values = c("cyan4", "orange")) +
  labs(x = latex2exp::TeX("$\\partial \\bar{T} / \\partial y$ change (K/km)"), 
       y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
       color = NULL, shape = NULL) +
  coord_cartesian(ylim = c(-4, 4)) +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.73, 0.1),
        legend.direction = "horizontal",
        legend.box.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        legend.margin = margin(-2, 2, 1, 2, unit = "mm"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA))

eke <- change |> 
  ggplot(aes(mean_eke, mean_deltat)) +
  geom_point(aes(color = region, shape = land), size = 2) +
  geom_smooth(aes(colour = region), method = "lm", se = FALSE) +
  scale_color_manual(values = c("cyan4", "orange")) +
  labs(x = latex2exp::TeX("EKE change ($m^2 s^{-2}$)"), 
       y = NULL,
       color = NULL, shape = NULL) +
  coord_cartesian(ylim = c(-4, 4)) +
  theme_minimal() +
  theme(legend.position = "none",
        # legend.position.inside = c(0.83, 0.08),
        # legend.direction = "horizontal",
        # legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        # legend.margin = margin(-3, 0, 0, 0, unit = "mm"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA))

plot_grid(gradt, eke, labels = c('a)', 'b)'), 
          label_x = 0.12, label_y = 0.98,
          label_fontface = "plain", 
          label_colour = "black",
          rel_widths = c(1, 1))
```


## Summary and Discussion

In this study, we use day-to-day falls in near-surface maximum temperature as a method to detect cold fronts and to evaluate changes in their frequency under future warming, in a set of 10 models from CMIP6. By using a specific percentile threshold, we ensure a consistent detection of strong temperature falls across different regions, improving the detection particularly over oceans where the temperature variability is lower. While the temperature drop method may misidentify other forms of temperature variability as fronts, previous studies have shown that this methodology provides a simple and robust method for characterising the frequency of fronts over midlatitude regions.

Our results show a general decrease in the frequency of fronts across the Northern Hemisphere, especially over oceans, and a poleward shift in frontal activity in the Southern Hemisphere. Over land in the Southern Hemisphere, notably in southern Australia and Africa, we find an increase in frontal frequency and intensity. These findings align with previous studies using more complex methods to identify fronts, strengthening confidence in our approach.

We find a positive correlation between changes in $\Delta T$ frequency and the meridional temperature gradient and transient eddy kinetic energy (EKE). 
This is consistent with a model for midlatitude temperature variability driven primarily by advection, and suggests that the large-scale baroclinic background state is an important control on front frequency in a warming climate. Further work could tease apart the individual effects of temperature gradients versus the intensity of the storm track, noting that these are linked, since changes in the mean thermodynamic structure affect how the storm track behaves [OGorman2010].
An important exception to the above correlations is that, in Southern Hemisphere land regions, front frequency increases despite a decrease in the EKE of the storm track. 
This may reflect the fact that storm track dynamics have less direct control over frontal systems, which could be more influenced by mesoscale processes, land–sea contrast, and regional thermodynamic effects.

One important factor not analyzed here is the role of sea surface temperature (SST) gradients in frontal development. @Parfitt2017 showed that the surface sensible heat flux can amplify a front, through diabatic frontogenesis  in regions with strong SST gradients, such as the Gulf Stream or Kuroshio Current. However, @Reeder2021 showed that atmospheric fronts along SST fronts are amplified primarily by adiabatic frontogenesis, i.e. the deformation of the wind field that help increase the equivalent potential temperature gradient. A projected weakening of SST gradients under climate change and as a consequence, the sensible heat flux and wind circulation, could partially explain the changes in frequency of fronts along the west coast of North America and Asia. Further analysis is needed to quantify the contribution of these mechanisms to projected changes.


## Acknowledgments

This work was supported by the Australian Research Council Centre of Excellence for Climate Extremes grant CE170100023. This research was undertaken with the assistance of resources from the National Computational Infrastructure (NCI Australia), an NCRIS enabled capability supported by the Australian Government. 
P.C. also acknowledges the use of the following open-source software projects that supported data analysis, visualization, and reproducible reporting: R (v4.3.1, @R), metR [@metR], rcdo [@rcdo], data.table [@data.table], ggplot2 [@ggplot2], patchwork [@patchwork], cowplot [@cowplot], ggside [@ggside], Quarto [@quarto].

**Should I also mention 21st century weather?**

## Open research

The CMIP6 data used in the manuscript is publicly available at the Earth System Grid  Federation [https://esgf-node.llnl.gov/projects/cmip6/](https://esgf-node.llnl.gov/projects/cmip6/). A version-controlled repository of the code used to create this analysis, including the necessary code to download the derived data can be found at <https://github.com/paocorrales/t-drop-trends>. The derived data that support the findings of this study are also openly available in Zenodo at http://doi.org/, version 0.xx. 


## References {.unnumbered}

:::{#refs}

:::