---
title: "Using rapid temperature falls to estimate future strong cold front frequency in CMIP6 climate projections"
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
    docx: default
author:
  - name: Paola B Corrales
    affiliations:
      - name: Monash University
        department: School of Earth, Atmosphere and Environment
        region: Melbourne
        country: Australia
      - name: ARC Centre of Excellence for Climate Extremes
        # department: ARC Centre of Excellence for Climate Extremes
        region: Melbourne
        country: Australia
      - name: Australian National University
        department: Research School of Earth Sciences
        region: Canberra
        country: Australia
      - name: ARC Centre of Excellence for the Weather of the 21st Century
        region: Melbourne
        country: Australia
    corresponding: true
    orcid: 0000-0003-1923-9129
    email: paola.corrales@monash.edu
  - name: Martin Singh
    affiliations:
      - name: Monash University
        department: School of Earth, Atmosphere and Environment
        region: Melbourne
        country: Australia
      - name: ARC Centre of Excellence for Climate Extremes
        region: Melbourne
        country: Australia
      - name: ARC Centre of Excellence for the Weather of the 21st Century
        region: Melbourne
        country: Australia
    orcid: 0000-0002-4584-0885
    email: martin.singh@monash.edu
  - name: Michael J. Reeder
    affiliations:
      - name: Monash University
        department: School of Earth, Atmosphere and Environment
        region: Melbourne
        country: Australia
      - name: ARC Centre of Excellence for Climate Extremes
        region: Melbourne
        country: Australia
    orcid: 000-0002-4583-5875
    email: michael.reeder@monash.edu
abstract: |
  We use day-to-day falls in daily-maximum surface air temperature to assess future changes in the frequency of strong cold fronts in a set of CMIP6 climate models under the SSP5-8.5 scenario. We find a general decrease in cold front frequency across the Northern Hemisphere, particularly over the oceans, and in the Southen Hemisphere a poleward shift in the region of highest frontal frequency, with increased frequency over midlatitude land regions. The frequency of strong temperature falls correlates positively with the meridional temperature gradient and transient eddy kinetic energy, implicating large-scale baroclinicity and changes in  the storm track influence the pattern of strong cold fronts. Over Southern Hemisphere land areas, this relationship weakens, suggesting the influence of regional thermodynamic processes.
plain-language-summary: |
  Cold fronts, are defined by large temperature falls and sometimes accompanied by rain and strong wind. They play a key role in the day-to-day weather in midlatitudes. In this study, we use a simple method based on changes in maximun temperature from one day to the next to identify strong cold fronts in climate simulations. In the future climate projections from ten global models, strong cold fronts are less frequent in much of the Northern Hemisphere, especially over the oceans. In the Southern Hemisphere, however, strong cold fronts in the projections shift closer to the South Pole and may become more frequent over land areas like southern Australia and southern Africa. These changes are linked to shifts in the temperature contrast between the equator and the poles and the paths taken by storm systems. 
keywords: []
keypoints: 
  - Day-to-day falls in maximum temperature are used to identify cold fronts in CMIP6 models using the historical and SSP5-8.5 simulations
  - In the NH the frequency of strong cold fronts decreases, while in the SH they shift poleward over the oceans and increase over land
  - Cold front frequency changes correlated positively with the meridional surface temperature gradient globally and with storm tracks in the NH
bibliography: [cold-fronts.bib, packages.bib]  
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
execute: 
  cache: true
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup

library(metR)
library(ggplot2)
library(data.table)
library(here)
library(cowplot)
library(ggside)
library(patchwork)

global_map <- rnaturalearth::ne_coastline()
models_in_gadi <- c("CMCC-ESM2", "EC-Earth3", "INM-CM4-8", "INM-CM5-0", 
                    "MPI-ESM1-2-HR", "NorESM2-MM", "EC-Earth3-CC", "EC-Earth3-Veg",
                    "EC-Earth3-Veg-LR", "GFDL-CM4")

file_list <- c(Sys.glob(here("data/eke/*hist*mean.nc")),
               Sys.glob(here("data/deltat/*hist*mean.nc")))

delta <- purrr::map(file_list, function(f) {
  
  message(f)
  meta <- unglue::unglue(basename(f), "{var}_{model}_{scenario}_mean.nc")
  
  
  if (meta[[1]][["var"]] == "eke") {
    var <- "ua"
  } else if (meta[[1]][["var"]] == "deltap97.5") {
    var <- "tasmax"
    meta[[1]][["var"]] <- "deltat"
  } 
  
  file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/*_", meta[[1]][["model"]], "_ssp585_mean.nc")))
  
  ReadNetCDF(f, vars = c(hist = var)) |> 
    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
    _[, let(model = meta[[1]][["model"]],
            var = meta[[1]][["var"]],
            change = ssp585 / hist,
            diff = ssp585 - hist,
            lat = round(lat, digits = 8),
            plev = NULL)] |> 
    _[]
  
}) |> rbindlist() 


ta <- purrr::map(Sys.glob(here("data/ta/*hist*mean.nc")), function(f) {
  
  message(f)
  meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_historical_mean.nc"))
  
  file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/", 
                                        meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585_mean.nc")))
  
  var <- meta[[1]][["var"]]
  
  ReadNetCDF(f, vars = c(hist = var), subset = list(plev = 85000)) |> 
    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), 
                                subset = list(plev = 85000),
                                out = "vector")[[1]])] |> 
    _[, let(model = meta[[1]][["model"]],
            var = var,
            lat = round(lat, digits = 8),
            hist = metR:::.derv(hist, lat)*(180/pi/6371000),
            ssp585 = metR:::.derv(ssp585, lat)*(180/pi/6371000)), by = .(lon)] |> 
    _[, let(change = ssp585 / hist,
            diff = ssp585 - hist,
            plev = NULL)]
  
}) |> rbindlist() 

data <- rbindlist(list(delta, ta)) |> 
  _[, let(land = MaskLand(lon, lat))] |>
  _[, let(land = factor(land, labels = c("Sea", "Land")))]

file_list <- c(Sys.glob(here("data/percentiles/*p97.5_1950-1979.nc")))

percentiles <- purrr::map(file_list, function(f) {
  
  message(f)
  meta <- unglue::unglue(basename(f), "{var}_{model}_{scenario}_{things}.nc")
  
  ReadNetCDF(f, vars = c(deltat = "tasmax")) |> 
    _[, let(model = meta[[1]][["model"]],
            lat = round(lat, digits = 8))] |> 
    _[]
  
}) |> rbindlist() 

```



## Introduction

Cold fronts are an important feature of weather in extratropical regions and their associated circulation contribute a substantial fraction of the precipitation in these regions [@Konstali2024; @Catto2013]. Cold fronts are at times linked to extreme events, such as heavy precipitation [@Catto2013] and severe wildfires in southern Australia [@Reeder2015] and other regions. Understanding how cold fronts, and especially strong cold fronts, change in a warming climate is therefore of importance for understanding the changing likelihood of extremes and fire weather.

Several computational methods have been developed to identify cold fronts using different characteristics. Some of these methods look for spatial gradients in temperature and humidity at different atmospheric levels [@Berry2011], others identify gradients in the wind field alone [@Simmonds2012] or a combination of dynamic and thermodynamic variables [@Thomas2019]. These methods have been developed for the data on a regular grid (e.g. outputs from numerical models or reanalysis) and their results depend on setting different thresholds, which may need to be calibrated for different grid resolutions. 


Here, we take the simplest approach, first introduced by @Reeder2015. We measure the occurrence of cold fronts using temporal changes in temperature at a fixed location. Specifically, we identify the passage of cold fronts with large day-to-day falls in near-surface daily-maximum temperature. 
This method has been shown to be suitable for the detection of strong cold fronts in southeastern Australia [@Cai2022; @Reeder2015], and can successfully identify cold fronts in data with different resolutions, including both station data and reanalyses [@King2024]. @King2024 developed a climatology of cold-front frequency in global climate models from phase 6 of the Coupled Model Intercomparison Project (CMIP6) which produced results comparable to other, more complex, cold-front detection algorithms. While the models tend to overestimate front frequency in polar regions and underestimate the occurrence of large temperature falls over midlatitude land regions compared with ERA5, they accurately represent the patterns of front frequency equatorward of about 60°.

As global temperatures continue to rise, understanding how this warming will affect the occurrence of cold fronts is crucial for projecting changes in extreme weather and fire risk. One of the few studies to address this question is @Catto2014, who identified fronts using an objective detection method based on temperature and humidity fields [@Berry2011]. In this framework, fronts are represented as line features, marking sharp gradients in thermodynamic variables. Using this approach in 18 CMIP5 models, they compared historical simulations (1980–2005) with projections under the Representative Concentration Pathway 8.5 scenario (2080–2100). They found an overall decrease in front frequency in the Northern Hemisphere, with a poleward shift in the latitude of maximum frequency and a significant decrease at high latitudes. The Southern Hemisphere also showed a poleward shift in the frequency maximum.

In contrast, @Konstali2024 identified fronts in CESM2-LE simulations using the gradient of equivalent potential temperature at 850 hPa. In this approach, frontal systems are represented as areas of strong gradients rather than discrete lines. Using this methodology, they found an increase in the frequency of fronts, which they suggest may be associated with fronts becoming stronger or more elongated in both hemispheres.

Motivated by the contrasting results of previous studies, we re-examine how the distribution of cold fronts change in projections from the CMIP6 ensemble, but we consider a simpler method of estimating changes in front frequency and intensity with cooling based on day-to-day falls in maximum surface air temperature.
This approach offers several advantages. First, the method's simplicity and weak resolution-dependence [as shown by @King2024] allows us to compare it with earlier studies using other methodologies. Second, focusing on local temperature falls provides a complementary perspective to Lagrangian approaches, as day-to-day temperature falls are the most direct impact of cold fronts on local weather.

Consistent with @Catto2014, we find a decrease in cold front frequency in the Northern Hemisphere and a poleward shift in the Southern Hemisphere. Seeking to develop a physical explanation for these changes, we first study the response of the meridional temperature gradient and its relationship with the change in front frequency. We also explore the relationship between changes in front frequency and transient eddy kinetic energy, as it is directly related to the storm tracks. Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@OGorman2010; @Tamarin2017; @Chemke2022], and these changes could also influence the location and intensity of future cold fronts.


## Data and Methods

We analyse 10 models from phase 6 of the Coupled Model Intercomparison Project (CMIP6, @Eyring2016), including the Historical experiment up to 2014 and the Shared Socioeconomic Pathways 5 (SSP5-8.5) from the Scenario Model Intercomparison Project (ScenarioMIP, @ONeill2016) from 2015 to 2100, as described in @tbl-models. We selected these models based on the data available for the  historical and scenario experiments stored on the Australian National Computational Infrastructure (NCI) replica data store.


```{r}
#| label: tbl-models
#| tbl-cap: "List of the coupled model intercomparison project phase 6 (CMIP6) climate model used in this analysis."
#| results: asis


readr::read_rds(here("data/cmip_file_list_scenario.rds")) |> 
  _[nominal_resolution == "100 km" & source_id %in% models_in_gadi, 
    .(source_id, institution_id, member_id)] |> 
  unique(by = "source_id") |> 
  _[order(source_id)] |> 
  _[, let(references = c("@Lovato2021; -@Lovato2021a", "@ConsortiumEC-Earth2019; -@ConsortiumEC-Earth2019a",
                         "@ConsortiumEC-Earth2020; -@ConsortiumEC-Earth2021", 
                         "@ConsortiumEC-Earth2019b; -@ConsortiumEC-Earth2019c",
                         "@ConsortiumEC-Earth2020a; -@ConsortiumEC-Earth2023",
                         "@Guo2018; -@Guo2018a", "@Volodin2019, @Volodin2019a",
                         "@Volodin2019b; @Volodin2019c", "@Jungclaus2019; @Schupfner2019",
                         "@Bentsen2019; @Bentsen2019a"))] |> 
  _[, let(source_id = paste0(source_id, "\n[", references, "]"))] |> 
  _[, .(source_id, institution_id, member_id)] |> 
  knitr::kable(col.names = c("Model", "Institution", "Ensamble member used"),
               caption = "List of models used for the analysis.",
               align = "lll",
               format = "markdown", format.args = list(zero.print = "-"))

```


### Front Identification

We identify falls in near-surface (2 meter) air temperature as a proxy for frontal passages by calculating the change in daily maximum temperature from one day to the next ($\Delta T$). $\Delta T$ is calculated for all grid points and days in the study period. @King2024 defined a passage of a strong cold front as $\Delta T \le -10^{\circ}C$. That is, a cold front passage is identified on day $i$ if the maximum temperature on day $i+1$ is at least 10 deg C lower than that of day $i$. While this value is reasonable for detecting strong falls in temperature over land, it may underestimate the frequency of fronts over the ocean due to the weaker temperature variability. For this reason, in this study, we use the 2.5$^{th}$ percentile of $\Delta T$ as a threshold for each grid point to identify strong falls in temperature, hereafter $\Delta T_{2.5} = \Delta T \le 2.5^{th}$ percentile. This percentile is chosen because the mean of the 2.5$^{th}$ percentile over all midlatitude land areas (between 25$^{\circ}$ to 55$^{\circ}$) is close to $-10^{\circ}C$, the threshold used by @King2024. The percentiles are calculated over the historical period between 1979 and 2014 for each model at each grid point. Additionally, the percentile values are also calculated over the same period for December-January-February (DJF) and June-July-August (JJA) for the seasonal analysis (see Supplemental Information). 

It is important to note that by using this threshold we detect frontal passages in exactly 2.5% of days. In regions of regular frontal passage the 2.5$^{th}$ percentile of $\Delta T$ generally corresponds to strong fronts, and our identification method produces reasonable results. However, outside these regions (e.g., in the tropics) we identify temperature falls that can be associated with midlatitude front-like features but also other processes. 



### Time-mean Meridional Temperature Gradient

We calculated the time-mean meridional temperature gradient ($\partial \bar{T}/ \partial y$) at each grid point using central finite differences for the temperature at 850 hPa. We then area-averaged $\partial \bar{T}/ \partial y$ over different regions, e.g.,  between 25$^{\circ}$ and 65$^{\circ}$ latitude for each hemisphere. This process produces the same results as if calculating the temperature difference between two latitudes but has the advantage of allowing the calculation of $\partial \bar{T}/ \partial y$ for regions over land and sea independently. 

### Storm Tracks Estimation {#sec-storm-track}

To estimate the storm track location we follow previous work [@OGorman2010; @Chemke2022] and use the high-pass filtered transient eddy kinetic energy (EKE) in the upper troposphere. For each model, we calculated the EKE as $EKE = 0.5(u'^2 + v'^2)$, where  $u'$ and $v'$ are obtained using a Fast Fourier Transform high-pass filter applied to daily horizontal wind at 250 hPa with a 6-day cutoff.  



## Results

@fig-trend a) shows the change in the frequency of $\Delta T_{2.5}$ between the yearly averaged $\Delta T_{2.5}$ frequency for last 20 years of the 21$^{st}$ century (SSP5-8.5 experiment) and
the last 20 years of the 20$^{th}$ century (historical experiment). @fig-trend b and c) show the zonal mean of $\Delta T_{2.5}$ frequency for each latitude over sea and over land respectively.

@fig-trend d) shows the $\Delta T$ values that define the 2.5$^{th}$ percentile. We found that the spatial pattern correlates with the climatology of cold fronts from @Berry2011 (their Figure 2a) and @King2024 (their Figure 6). This consistency supports our proposition that temperature falls identified here are directly related to cold front passages.

In terms of projected changes, in the Northern Hemisphere over land, there is a robust decrease in $\Delta T_{2.5}$, particularly north of 40$^{\circ}$ latitude, except in western and central Europe where the multi-model mean shows an increase but the signal is not robust across the ensemble. A similar pattern of changes in front frequency was also found by @Catto2014 using CMIP5 and a different methodology to identify fronts, reinforcing the reliability of the methodology used in this analysis. Over the oceans, front frequency also tends to decrease over the North Atlantic, which shows a multi-model mean increase in front frequency west of continental Europe. This local increase may be interpreted as a shift of front frequency toward the pole in this sector, although it does not appear consistently across the ensemble.

In the Southern Hemisphere over land, the change is generally positive with an increase of 5 strong temperature falls per year in some areas. Other authors have also found an increase in the frequency of fronts in the south and east regions of Africa and Australia [@Catto2014; @Konstali2024]. 

Changes in $\Delta T_{2.5}$ frequency over the Southern Hemisphere ocean regions are less robust across the ensemble, but the ensemble mean shows a tripolar pattern, with increases in a band centered around 40$^{\circ}$-50$^{\circ}$ S (depending on the longitude) and decreases poleward and equatorward. 
This pattern indicates that the areas of higher front frequency shift poleward in the future scenario. In the Indian ocean, the pattern in more complicated, with a second band of increased frequency in the subtropics is most pronounced in the Agulas region. 

The analysis of the change in frequency of $\Delta T_{2.5}$ by season in DJF and JJA (Figure S1 a) shows similar spatial patterns and the same sign of the change when compared with the change per year in @fig-trend a). The main exception is western Europe, where we find an increase in front frequency during winter but a decrease in summer.

```{r}
#| label: fig-trend
#| fig-width: 8
#| fig-height: 8
#| fig.cap: "a) Multimodel mean change in the frequency of $\\Delta T_{2.5}$ (days/year, SSP5-8.5 2080-2100 minus Historical 1980-2000) and zonal mean change over b) land and c) sea, each grey line represents a model listed in Table 1 and a black line represents the multimodel mean. In a) dotted areas show where at least 90% of the models agree in the sign of change. d) $\\Delta T$ values associated with the 2.5$^{th}$ percentile and zonal mean over e) land and f) sea."

topo <- rcdo::cdo_topo(grid = "r360x180") |> 
  rcdo::cdo_execute(options = "-f nc") |> 
  ReadNetCDF()

sig <- data |> 
  _[(model %in% models_in_gadi) & var == "deltat"] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, .(sig = sum(sign(deltat), na.rm = TRUE)), by = .(lon, lat)] 

map_trend <- data |> 
  _[(model %in% models_in_gadi) & is.finite(change)] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, .(mean_dt = mean(deltat, na.rm = TRUE)), by = .(lon, lat)] |> 
  ggplot(aes(ConvertLongitude(lon), lat)) +
  geom_contour_fill(aes(z = mean_dt, fill = after_stat(level)),
                    breaks = c(-Inf, seq(-8, 8, 0.5), Inf)) +
  scale_fill_divergent_discretised(name = "days/year", mid = "#f7f7f7",
                                   guide = guide_colorbar(barheight = 0.5),
                                   labels = function(x) JumpBy(x, 2, fill = "")) +
  geom_sf(data = global_map, inherit.aes = FALSE, fill = NA, linewidth = 0.2) +
  geom_point(data = sig[is.cross(lon, lat, skip = 1) & abs(sig) >= 9], alpha = 0.3, shape = 4, size = 0.1) +
  scale_y_latitude(breaks = seq(-80, 80 , 20)) +
  coord_sf(expand = FALSE, xlim = c(-179, 180)) +
  tagger::tag_facets(tag_pool = c("a")) +
  labs(x = NULL, y = NULL, fill = "days/year") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.frame = ggplot2::element_rect(color = "black", linewidth = 0.4),
        legend.key.width = grid::unit(1, 'null'),
        legend.margin = unit(c(-1, 0, -1, 0), units = "cm"), 
        plot.margin = unit(c(-0.5, 0 , -0.5, 0), units = "cm"),
        tagger.panel.tag.text = element_text(size = 11),
        tagger.panel.tag.background = element_rect(fill = "white", color = "white")) 

annotation <- data.frame(ymin = -66, ymax = -56, xmin = -Inf, xmax = Inf, land = "Land")

mean_trend <- data |> 
  _[(model %in% models_in_gadi) & var == "deltat"] |>
  dcast(lat + lon + model ~ var, value.var = "diff") |> 
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, .(mean_dt = mean(deltat, na.rm = TRUE)), by = .(lat, land, model)] |> 
  ggplot(aes(mean_dt, lat)) +
  geom_vline(xintercept = 0, color = "grey50") +
  geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
  stat_summary(orientation = "y", geom = "line") +
  scale_y_latitude(breaks = seq(-80, 80 , 20), position = "right") +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#fbfbfb",
            data = annotation,
            inherit.aes = FALSE) +
  facet_wrap(~land) +
  tagger::tag_facets(tag_pool = c("b", "c"), position = "tr") +
  labs(x = "days/year") +
  coord_fixed(ratio = 1/2.8) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        panel.grid.minor = element_line(linewidth = 0.1),
        strip.text.x = element_text(size = 12),
        plot.margin = unit(c(0.15, 0, 1.75, 0.25), units = "cm"),
        tagger.panel.tag.text = element_text(size = 12),
        tagger.panel.tag.background = element_rect(fill = "white", color = "white")) 

front <- plot_grid(map_trend, mean_trend,
          rel_widths = c(2.25, 1)) +
  theme(plot.margin = margin(t = 0, b = -1.5, unit = "cm"))

map_perc <- percentiles |> 
  _[, .(mean_delta = mean(deltat)), by = .(lon, lat)] |> 
  ggplot(aes(ConvertLongitude(lon), lat)) +
  geom_contour_fill(aes(z = -mean_delta, fill = after_stat(level)),
                    breaks = c(seq(0, 16, 1), Inf)) +
  scale_fill_viridis_d(name = latex2exp::TeX("$\\Delta T$ (K)"),
                       guide = guide_colorsteps(barheight = 0.5),
                       direction = -1) +
  geom_sf(data = global_map, inherit.aes = FALSE, fill = NA, linewidth = 0.2) +
  scale_y_latitude(breaks = seq(-80, 80 , 20)) +
  coord_sf(expand = FALSE, xlim = c(-179, 180)) +
  labs(x = NULL, y = NULL) +
  tagger::tag_facets(tag_pool = c("d")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.frame = ggplot2::element_rect(color = "black", linewidth = 0.4),
        legend.key.width = grid::unit(1, 'null'),
        legend.margin = margin(0, 0, 0, 0, unit = "cm"), 
        plot.margin = unit(c(0, 0 , -1, 0), units = "cm"),
        tagger.panel.tag.text = element_text(size = 11),
        tagger.panel.tag.background = element_rect(fill = "white", color = "white")) 

mean_perc <- percentiles |> 
  _[(model %in% models_in_gadi)] |>
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, .(mean_dt = mean(deltat, na.rm = TRUE)), by = .(lat, land, model)] |> 
  ggplot(aes(mean_dt, lat)) +
  geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
  stat_summary(orientation = "y", geom = "line") +
  scale_x_continuous(breaks = seq(-12, -0, 4)) +
  scale_y_latitude(breaks = seq(-80, 80 , 20), position = "right") +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#fbfbfb",
            data = annotation,
            inherit.aes = FALSE) +
  facet_wrap(~land, labeller = labeller(land = c(Land = " ", Sea = " "))) +
  tagger::tag_facets(tag_pool = c("e", "f"), position = "tr") +
  labs(x = latex2exp::TeX("$\\Delta T$ (K)")) +
  coord_fixed(ratio = 1/5) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        panel.grid.minor = element_line(linewidth = 0.1),
        strip.text.x = element_text(size = 12),
        plot.margin = unit(c(1, 0, 1.5, 0.25), units = "cm"),
        tagger.panel.tag.text = element_text(size = 12),
        tagger.panel.tag.background = element_rect(fill = "white", color = "white")) 

perc <- plot_grid(map_perc, mean_perc,
          rel_widths = c(2.25, 1)) +
  theme(plot.margin = margin(t = -1.5, unit = "cm"))

plot_grid(front, perc, ncol = 1, rel_heights = c(1.3, 1))
```

@fig-serie shows ensemble-mean time series of the zonal-mean frequency of strong temperature falls for different latitude bands in each hemisphere. Consistent with the map in @fig-trend, the trends in frequency are negative over ocean in the Northern Hemisphere, but positive over land equatorward of 50$^{\circ}$ N. Over the Southern Hemisphere, trends are generally positive over land, with changes over ocean generally consistent with a poleward shift of fronts.

While there is considerable variation in these trends between models (see boxplots on right of @fig-serie), particularly at more poleward locations, for most of the zonal mean trends, there is consistency in the sign of the trend across models.


```{r}
#| label: fig-serie
#| fig-width: 8
#| fig-height: 5
#| fig.cap: "Multimodel yearly mean frequency of $\\Delta T_{2.5}$ for the historical period (1900-2014) and SSP5-8.5 (2015-2100) over land and sea for different 2° wide latitude bands centered at 30, 40 and 50°. The grey line represents the climatological frecuency for 1979-2014, equal to 9.125 $\\Delta T_{2.5}$/year. On the right of each panel box plots of the intermodel distribution is shown for the period 2080-2100. In the boxplots the lower and upper hinges correspond to the first and third quartiles and the upper (lower) whisker extends to the largest (smallest) value no further than 1.5 * inter-quartile range. Outside the whiskers dots represents outliers."

serie <- purrr::map(Sys.glob(here("data/series/serie_freq_deltap97.5_*_100km.nc")), function(f) {
  
  meta <- unglue::unglue(basename(f), "serie_freq_{threshold}_{scenario}_100km.nc")
  ReadNetCDF(f, vars = c(mean_freq = "tasmax")) |> 
    _[, let(year = year(time),
            scenario = meta[[1]][["scenario"]])] |> 
    _[, let(time = NULL)]
  
}) |> rbindlist() |> 
  _[, let(scenario = factor(scenario, labels = c("Historical", "SSP5-8.5")))]

model_spread <- data |> 
  _[var == "deltat" & model %in% models_in_gadi] |> 
  _[, let(serie = fcase(lat %between% c(48, 50), "50° - NH",
                        lat %between% c(38, 40), "40° - NH",
                        lat %between% c(28, 30), "30° - NH", 
                        lat %between% c(-50, -48), "50° - SH",
                        lat %between% c(-40, -38), "40° - SH",
                        lat %between% c(-30, -28), "30° - SH"))] |>
  _[!is.na(serie)] |>
  _[, .(mean_freq = weighted.mean(ssp585, cos(lat*pi/180))), by = .(serie, land, model)] |> 
  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
  setDT() |> 
  _[, let(scenario = "SSP5-8.5")] |> 
  _[!(land == "Land" & hemisphere == "SH" & lat == "50°")]

serie_regions <- serie |>
  _[!(year == 2014 & scenario != "historical")] |>
  _[, let(land = MaskLand(lon, lat))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
  _[, let(serie = fcase(lat %between% c(48, 50), "50° - NH",
                        lat %between% c(38, 40), "40° - NH",
                        lat %between% c(28, 30), "30° - NH", 
                        lat %between% c(-50, -48), "50° - SH",
                        lat %between% c(-40, -38), "40° - SH",
                        lat %between% c(-30, -28), "30° - SH"))] |>
  _[!is.na(serie)] |> 
  _[, .(mean_freq = weighted.mean(mean_freq, cos(lat*pi/180))), by = .(land, serie, year, scenario)] |> 
  _[!(land == "Land" & serie == "50° - SH")]

serie_regions |> 
  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
  setDT() |> 
  _[year >= 1900] |> 
  ggplot(aes(year, mean_freq, color = lat, group = interaction(scenario, lat))) +
  geom_line(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_segment(x = 1900, xend = 2100, y = 9.125, yend = 9.125, color = "grey50") +
  geom_ysideboxplot(data = model_spread, aes(x = lat), orientation = "x") +
  scale_color_manual(values = c("#004C73", "#CC3311", "#ee8a33")) + 
  facet_grid(land~hemisphere, scale = "free") +
  labs(x = "Year", y = latex2exp::TeX("$\\Delta T_{2.5}$ frequency (days/year)"), color = "Latitude") +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.78, 0.07),
        legend.direction = "horizontal",
        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        strip.text = element_text(size = 12),
        ggside.axis.text.x = element_blank(),
        ggside.panel.scale = 0.2)

```

To explore the distribution of the change in temperature falls of different intensity, @fig-distribution shows the change in the frequency of $\Delta T$ using different percentiles, from 10$^{th}$ to 0.1$^{th}$, as thresholds to identify the falls in temperature. The percentile values are again calculated from the historical period between 1979 and 2014 for each grid point and model as in @fig-trend. 

@fig-distribution shows that, in most cases, the sign of the change is the same across all percentile thresholds, meaning that there is a projected increase or decrease in the frequency of $\Delta T$ regardless of the intensity of the temperature fall. The exceptions are the Southern Hemisphere over sea and the Northern Hemisphere over land in the 40 $^{\circ}$ latitude band, where there is a modest increase in strong temperature falls but a small decrease in weaker falls. 

Moreover the change in the frequency of $\Delta T$ is greater at lower percentile thresholds, which identify more extreme temperature falls. For example, in the Northern Hemisphere over the sea, the 0.1$^{th}$ percentile threshold shows that there will be less than half as many falls in temperature in the future over 50$^{\circ}$ N, with the opposite expected over land in the Southern Hemisphere. 
However, lower percentile thresholds are associated with rarer events, meaning fewer occurrences per year. These results suggest that regions projected to experience more temperature falls in the future, such as land areas in midlatitudes in the Southern Hemisphere, will also experience larger $\Delta T$, whereas regions with a decreasing frequency of falls will see smaller falls.


```{r}
#| label: fig-distribution
#| fig-width: 8
#| fig-height: 5
#| fig.cap: "Fractional change in the frequency of strong $\\Delta T$ averaged between the years 2080-2100 under the SSP5-8.5 scenario compared to that averaged between 1980-2000 in the historical scenario, for different thresholds defined by their percentile value for each grid point and model calculated over the historical period (1979-2014)."

file_list <- Sys.glob(here(paste0("data/series/serie_freq_deltap9*", c("historical", "ssp585"), "*")))

dist <- purrr::map(file_list, function(f) {
  message(f)
  
  meta <- unglue::unglue(basename(f), "serie_freq_deltap{percentile}_{scenario}_100km.nc")
  
  if (meta[[1]][["scenario"]] == "historical") {
    
    ReadNetCDF(f, vars = "tasmax", subset = list(time = c("1980-01-01", "2000-12-31"))) |> 
      _[, .(mean_freq = mean(tasmax)), by = .(lon, lat)] |> 
      _[, let(percentile = meta[[1]][["percentile"]],
              scenario = meta[[1]][["scenario"]])]
  } else {
    
    ReadNetCDF(f, vars = "tasmax", subset = list(time = c("2080-01-01", "2100-12-31"))) |> 
      _[, .(mean_freq = mean(tasmax)), by = .(lon, lat)] |> 
      _[, let(percentile = meta[[1]][["percentile"]],
              scenario = meta[[1]][["scenario"]])]
  }
  
}) |> rbindlist() |> 
  _[, let(land = MaskLand(lon, lat),
          region = fcase(lat %between% c(25, 55), "NH",
                         lat %between% c(-55, -25), "SH",
                         default = NA))] |> 
  _[, let(land = factor(land, labels = c("Sea", "Land")))] 

dist |> 
  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50° - NH",
                        lat %between% c(38, 40) & land == "Sea", "40° - NH",
                        lat %between% c(28, 30) & land == "Sea", "30° - NH", 
                        lat %between% c(48, 50) & land == "Land", "50° - NH",
                        lat %between% c(38, 40) & land == "Land", "40° - NH",
                        lat %between% c(28, 30) & land == "Land", "30° - NH",
                        lat %between% c(-50, -48) & land == "Land", "50° - SH",
                        lat %between% c(-40, -38) & land == "Land", "40° - SH",
                        lat %between% c(-30, -28) & land == "Land", "30° - SH",
                        lat %between% c(-50, -48) & land == "Sea", "50° - SH",
                        lat %between% c(-40, -38) & land == "Sea", "40° - SH",
                        lat %between% c(-30, -28) & land == "Sea", "30° - SH"))] |>
  _[!is.na(serie)] |> 
  _[, .(mean_freq = weighted.mean(mean_freq, cos(lat*pi/180))), by = .(land, serie, percentile, scenario)] |> 
  dcast(land + serie + percentile ~ scenario, value.var = "mean_freq") |> 
  _[, let(historical = fifelse(serie  == "50° - SH" & land == "Land", 0, historical),
          ssp585 = fifelse(serie  == "50° - SH" & land == "Land", 0, ssp585))] |> 
  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
  setDT() |> 
  ggplot(aes(factor(round(100-as.numeric(percentile), digits = 1)), ssp585 / historical)) +
  geom_col(aes(fill = lat), position = "dodge") +
  scale_fill_manual(values = c("#004C73", "#CC3311", "#ee8a33")) +
  scale_y_log10() +
  facet_grid(land~hemisphere) +
  labs(x = "Percentile", y = latex2exp::TeX("Fractional change of $\\Delta T_{2.5}$"),
       fill = "Latitude") +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.8, 0.07),
        legend.direction = "horizontal",
        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA),
        strip.text = element_text(size = 12))


```

### Mechanisms

We now investigate the mechanisms driving the changes in $\Delta T_{2.5}$ frequency described above. We take a simple perspective on midlatitude temperature variability based on advection as a key driver [@Schneider2015]. According to this view, changes in midlatitude temperature variability are closely linked to changes in large-scale temperature gradients and to the strength of mixing by geostrophic turbulence. Motivated by this perspective, we explore relationships between changes in front frequency and changes in the large-scale meridional temperature gradient ($\partial \bar{T} / \partial y$) and EKE within the midlatitude storm tracks.

@fig-eke a) shows that, in both hemispheres, there is a strong correlation between increases in front frequency and changes in the large-scale meridional surface temperature gradient $\partial \bar{T} / \partial y$ across models. Here, $\partial \bar{T} / \partial y$ is taken as the time and zonal-mean meridional temperature gradient averaged between 25$^{\circ}$ and 65$^{\circ}$ in each hemisphere over either land or ocean.
As previously noted by @Catto2014, reductions in front frequency in the Northern Hemisphere are associated with a decrease in the the temperature gradient near the surface, while opposite changes in surface temperature gradients in the Southern Hemisphere drive increases is front frequency there. Here, we find that this relationship holds within the model ensemble; larger increases or decreases in  $\partial \bar{T} / \partial y$ in a given model are associated with larger projected increases or decreases in $\Delta T$, a relationship that is particularly strong in the Northern Hemisphere. 


There is a similar relationship across the ensemble between changes in the frequency of $\Delta T_{2.5}$ and changes in the energy of the storm track. Here we estimate the storm track energy by taking the high-pass filtered eddy kinetic energy at 250 hPa (see @sec-storm-track) and integrating over the 25$^{\circ}$ and 65$^{\circ}$ latitudes in each hemisphere. Models with projected decreases in EKE tend to project larger decreases in the frequency of $\Delta T_{2.5}$. 
The exception is the Southern Hemisphere over land where all models project weak changes in EKE and large increases in the frequency of $\Delta T_{2.5}$. The reason for this difference is unclear, but could suggest the importance of local land processes in amplifying fronts over land in some regions of the Southern Hemisphere [@Muir2010]. Moreover, most of the land affected by fronts in the Southern Hemisphere is north of the mean storm track, and thus the storm track has only a weak effect on temperature variability there.
Supporting this, @Harvey2014 found that the storm track response in the Southern Hemisphere during summer is associated with the change in the low level $\partial \bar{T} / \partial y$ mainly poleward of 65$^{\circ}$ S, i.e. over the southern oceans. 

In agreement with the previous literature on seasonal changes in the midlatitude storm track [@Priestley2022; @OGorman2010], we find a general increase in EKE during summer and a decrease in winter in the Northern Hemisphere. There is also a general positive change in the Southern Hemisphere in both seasons associated with a poleward shift of the storm tracks (Figure S2 c-d). Given these seasonal asymmetries, it is important to note that the change in EKE averaged over the entire year (@fig-eke b) is dominated by the winter season in each hemisphere. 

Notwithstanding the above caveats, our results suggest that changes in the mean meridional temperature gradient and storm track intensity play an important role in modulating changes in front frequency in a warming climate. Indeed, differences in how these quantities change in projections appear to account for the differences in front frequency changes between hemispheres. However, our results do not provide strong evidence for the relative importance of the two factors: aside from over Southern Hemisphere, both factors correlate with the $\Delta T_{2.5}$ frequency across the model ensemble and both are theoretically associated with changes in temperature variability.


```{r}
#| label: fig-eke
#| fig-width: 8
#| fig-height: 4
#| fig.cap: "Mean change in the mean $\\Delta T_{2.5}$ frequency and a) mean merdional temperature gradient $\\partial \\bar{T} / \\partial y$ and b) the EKE of the storm track for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model. Values are calculated for the Northern (25$^{\\circ}$ to 65$^{\\circ}$) and Southern (-25$^{\\circ}$ to -65$^{\\circ}$) Hemispheres and over sea (circles) and land (triangles). Each solid line corresponds to a linear regression for each hemisphere."

change <- data |> 
  _[(model %in% models_in_gadi)] |>
  dcast(lat + lon + model + land ~ var, value.var = "diff") |> 
  
  _[, let(region = fcase(lat %between% c(25, 65), "NH",
                         lat %between% c(-65, -25), "SH",
                         default = NA))] |> 
  _[, .(mean_eke = weighted.mean(eke, w = cos(lat*pi/180), na.rm = TRUE),
        mean_dt_dy = weighted.mean(ta, w = cos(lat*pi/180), na.rm = TRUE),
        mean_deltat = weighted.mean(deltat, w = cos(lat*pi/180), na.rm = TRUE)), by = .(region, land, model)] |> 
  _[, let(mean_dt_dy = fifelse(region == "NH", -mean_dt_dy, mean_dt_dy))] |> 
  na.omit()

r2_dt <- change |> 
  _[, .(r2 = round(cor(mean_deltat, mean_dt_dy)^ 2, digits = 2)), by = .(region, land)]

gradt <- r2_dt[change, on = .(region, land)] |> 
  ggplot(aes(mean_dt_dy*1000, mean_deltat, colour = region, group = interaction(region, land))) +
  geom_point(aes(color = region, shape = land), size = 2) +
  geomtextpath::geom_textsmooth(aes(label = r2, colour = region, group = interaction(region, land)),
                                method = "lm",
                                linewidth = 1,
                                hjust = 0.9,
                                vjust = 1.3,
                                size = 3,
                                gap = FALSE,
                                show.legend = FALSE) +
  scale_color_manual(values = c("cyan4", "orange")) +
  labs(x = latex2exp::TeX("$\\partial \\bar{T} / \\partial y$ change (K/km)"), 
       y = latex2exp::TeX("$\\Delta T_{2.5}$ frequency change (days/year)"),
       color = NULL, shape = NULL) +
  coord_cartesian(ylim = c(-4, 4)) +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.73, 0.1),
        legend.direction = "horizontal",
        legend.box.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
        legend.margin = margin(-2, 2, 1, 2, unit = "mm"),
        panel.background = element_rect(fill = "#fbfbfb", color = NA))


r2_dt <- change |> 
  _[, .(r2 = round(cor(mean_deltat, mean_eke)^ 2, digits = 2)), by = .(region, land)]

eke <- r2_dt[change, on = .(region, land)] |> 
  ggplot(aes(mean_eke, mean_deltat)) +
  geom_point(aes(color = region, shape = land), size = 2) +
  geomtextpath::geom_textsmooth(aes(label = r2, colour = region, group = interaction(region, land)),
                                method = "lm",
                                linewidth = 1,
                                hjust = 0.9,
                                vjust = 1.3,
                                size = 3,
                                gap = FALSE) +
  scale_color_manual(values = c("cyan4", "orange")) +
  labs(x = latex2exp::TeX("EKE change ($m^2 s^{-2}$)"), 
       y = NULL,
       color = NULL, shape = NULL) +
  coord_cartesian(ylim = c(-4, 4)) +
  theme_minimal() +
  theme(legend.position = "none",
        panel.background = element_rect(fill = "#fbfbfb", color = NA))

plot_grid(gradt, eke, labels = c('a)', 'b)'), 
          label_x = 0.12, label_y = 0.98,
          label_fontface = "plain", 
          label_colour = "black",
          rel_widths = c(1, 1))
```


## Summary and Discussion

In this study, we use day-to-day falls in near-surface maximum temperature to detect cold fronts, and we evaluate changes in their frequency under future warming in a set of 10 models from CMIP6. By using a specific percentile threshold, we ensure a consistent detection of strong temperature falls across different regions, improving the detection particularly over oceans where the temperature variability is lower. While the temperature fall method may occasionally misidentify other forms of temperature variability as fronts, previous studies have shown that this methodology provides a simple and robust method for characterising the frequency of fronts over midlatitude regions [@King2024].

Our results show a general decrease in the frequency of fronts across the Northern Hemisphere, especially over oceans, and a poleward shift in frontal activity in the Southern Hemisphere. Over land in the Southern Hemisphere, notably in southern Australia and Africa, we find an increase in frontal frequency and intensity. These findings align with previous studies using more complex methods to identify fronts, strengthening confidence in our approach.

We find a positive correlation between changes in $\Delta T_{2.5}$ frequency and the meridional temperature gradient and transient eddy kinetic energy (EKE). 
This is consistent with a model for midlatitude temperature variability driven primarily by advection, and suggests that the large-scale baroclinic background state is an important control on front frequency in a warming climate. Further work could tease apart the individual effects of temperature gradients versus the intensity of the storm track, noting that these are linked, since changes in the mean thermodynamic structure affect how the storm track behaves [@OGorman2010].
An important exception to the above correlations is that, in Southern Hemisphere land regions, front frequency increases despite a decrease in the EKE of the storm track. 
This may reflect the fact that storm track dynamics have less direct control over frontal systems in this region, which could be more influenced by mesoscale processes, land–sea contrast, and regional thermodynamic effects.

There are other factors not analyzed here that could contribute to the projected changes in frontal activity. For example, sea surface temperature (SST) gradients can influence the development and intensity of atmospheric fronts through air–sea interactions. @Parfitt2017 suggested that surface sensible heat fluxes associated with strong SST gradients, such as those found in the Gulf Stream or Kuroshio Current regions, can amplify atmospheric fronts through diabatic frontogenesis, although @Reeder2021 showed that the effect was only significant when the fronts were weak. In contrast, @Reeder2021 showed that strong atmospheric fronts along SST fronts are amplified primarily by adiabatic frontogenesis (i.e. the deformation of the wind field).

More generally, frontal cyclogenesis is directly affected by the wind and temperature fields. Consequently, projected changes in frontal frequency may also reflect modifications in the large-scale circulation, such as shifts in the background flow, changes in horizontal wind shear, or variations in baroclinicity. Further analysis is needed to quantify the contribution of these mechanisms to projected changes.

Finally, while this study focuses on cold fronts, in some regions warm fronts also play an important role in the distribution of extreme precipitation [@Catto2013]. We explored extending the methodology to warm fronts by considering day-to-day increases in maximum temperature. However, unlike cold fronts, warm fronts are not associated with a sufficiently distinct temperature signal to allow a simple threshold-based identification without also including changes associated with other local thermodynamic processes. Developing a simple and robust method for identifying warm fronts would extend the present framework and provide further insight into projected changes in frontal activity under climate change.


## Acknowledgments

This research was supported by the Australian Research Council Centre of Excellence for Climate Extremes (CE170100023), the Australian Research Council Centre of Excellence for the Weather of the 21st Century (CE230100012), with the assistance of resources from the National Computational Infrastructure (NCI Australia), an NCRIS enabled capability supported by the Australian Government. 
P.C. also acknowledges the use of the following open-source software projects that supported data analysis, visualization, and reproducible reporting: R (v4.3.1, @R), metR [@metR], rcdo [@rcdo], data.table [@data.table], ggplot2 [@ggplot2], patchwork [@patchwork], cowplot [@cowplot], ggside [@ggside], and Quarto [@quarto].

## Open research

The CMIP6 data used in the manuscript is publicly available at the Earth System Grid  Federation [https://esgf-node.llnl.gov/projects/cmip6/](https://esgf-node.llnl.gov/projects/cmip6/). A version-controlled repository of the code used to create this analysis, including the necessary code to download the derived data can be found at <https://doi.org/10.5281/zenodo.21713780>. The derived data that support the findings of this study are also openly available in Zenodo at <https://doi.org/10.5281/zenodo.21713532>. 


## References {.unnumbered}

:::{#refs}

:::